Dish: Distribution-oblivious Shell Scripting

### 2019-11-22 pldi/draft.tex
Pass of 5, 5.1, 5.2

\section{Implementation}
\label{impl}
% \kk{The writing in this section is wandering around without a clear
%   purpose :( }
% 
% \kk{It might make sense to merge this section with the one below with
%   the name: From Shell to IR and back}

@@

In order to address this issue, we introduce the notion of
distributable program regions, and we design a translation pass that
given the abstract syntax tree of a shell script identifies
distributable regions and translates each of them to a dataflow
graph. The AST of the shell script is produced using a POSIX compliant
parser \kk{Reference libdash github. This las sentence should probably
  go to evaluation}.

@@

%% \kk{I wanted this to be purely about the regions, and a next
%%   subsection be about the analysis, but the separation line is very
%%   blurry, so they are now in the same subsection.}

\kk{The goal of this subsection is to explain that Dish does a safe
  but effective analysis to identify which parts of the scripts to
  parallelize.}

@@

Notice that the structure of a shell program clearly identifies
components that can be executed concurrently, and barriers that
enforce synchronization. More precisely, \ttt{cat f1 f2} and
\ttt{grep ``foo''} would already execute concurrently in the
standard shell, while \ttt{sort f3} waits for their completion
before being executed. Both \ttt{cat f1 f2 | grep ``foo'' > f3} and
\ttt{sort f3} are maximal distributable regions, in the sense that
they cannot be extended beyond \ttt{\&\&}.

Concretely, a distributable region cannot permeate through the
following barrier constructs: ;, \&\&, \ttt{||}, function
definitions, command substitutions, and control flow primitives. On
the other hand, the pipe operator \ttt{|} and the parallel
composition operator \& enable parallelization, and extend
distributable regions. Essentially, distributed regions correspond to
subsections of the program that would be executed by different shell
processes.

\kk{I don't know whether I should mention the following: While these
  control flow constructs can be parallelized and there has been
  research on it, it is orthogonal to our work, and can be
  incorporated as future work.}

@@

Dish performs a depth first search pass on the AST of the given shell
program, extending the distributable regions from the bottom up,
translating their independent components to dataflow graph
nodes, until a barrier construct is reached. Note that barrier
constructs can be contained in a dataflow graph, as long as they don't
extend beyong a single node. An illustrative example follows:

@@

In the above example\kk{Can we make example not floating?}, while
\ttt{sort f1 > g1}, \ttt{sort f2 > g2}, and \ttt{sort -m g1
  g2} are all in different dataflow graphs. The dataflow graph that
corresponds to the whole program, has two nodes, \ttt{sort f1 > g1
  \&\& sort f2 > g2} and \ttt{sort -m g1 g2}.

In order to identify possible opportunities for parallelism, the
translation pass also identifies the category of each node, together
with its input and output, using the rules described in \ref{}. Some
interesting cases of the analysis algorithm are shown in pseudocode
below:

@@

%% \kk{Explain how the three above cases work, mentioning what does
%%   pipe-graphs and safe do. The important point to get across is to
%%   explain why we do the ``conservative checks'' in safe, and how they
%%   preserve safety.}

Whenever the algorithm encounters a pipe node, it recursively
translates its subcomponents, and then merges them in a dataflow graph
by connecting the output of the first with the input of the second,
\etc. Before the different dataflow graphs are connected, the
algorithm checks that at most one node in the graph writes at every
file, and at most one node reads from a file that another node writes
to. This is important, in order to avoid inconsistent behaviour after
distributing nodes in the dataflow graph \kk{Does this need an
  example?}. Whenever the algorithm encounter an \ttt{\&\&} node,
it recursively translates its subcomponents (similarly to the pipe
node), but then doesn't merge them in a large dataflow graph. Finally,
whenever the algorithm encounters a command, it first checks whether
it is safe to possibly parallelize this command. A command is
considered to be safe, if it does not make any modification to the
shell state (e.g. changing the current directory, modifying
environment variables), and if it does not read from a
\todo{read-once} file, like a fifo pipe. \kk{Mention: Note that an
  unsafe command can still be safely added to the dataflow graph as a
  single node, since this doesn't introduce any new
  behaviours. However it cannot be parallelized further than that.} If
the command is considered safe, its category is identified and a node
containing that command is created and added to a singleton dataflow
graph.

It is important to note that due to the highly dynamic nature of the
shell, the safety properties that were mentioned above cannot be
soundly checked statically. For example in the first case, two
unexpaned strings could refer to the same file, but the translation
pass cannot infer that. Hence, in both cases, the checks were
approximated using a \todo{best-effort} method. \kk{Maybe mention that
  as future work, we want to tighly integrate with the shell shtepper
  that will allow to maximally expand everything before making this
  decision.}

@@

\heading{Limitations}
Our is implementation is limited in many ways, so as to succeed in proving the key hypothesis---that scaling out shell pipelines can be automated and correct--- despite \sys being a research prototype.

@@

\sys is conservative in the shell subsets that it handles, so that it doesn't introduce any unsafe parallelism.
The point of this work was not to be able to handle a complete distributable
subset of the shell, but rather use significant part of it to demonstrate
performance benefits.

\heading{Future Work}
% \label{limitation}
% \item No cycles (multiple commands writing and reading from the same file)
There are several worthwhile directions for future work.
One direction would be a formal study of \sys and specifically a proof that the analysis~\sx{dataflow} does indeed return distributable subtrees.
Recent work on formalizing the semantics of the POSIX shell~\cite{smoosh:20} make such an attempt possible.

>>>>

\section{Implementation}
\label{impl}

@@

In order to address this issue, we \todo{introduce} the notion of
distributable program regions, and we design a translation pass that
given the abstract syntax tree of a shell script identifies
distributable regions and translates each of them to a dataflow
graph. The AST of the shell script is produced using a POSIX compliant
parser~\cite{libdash}.

@@

%% \kk{The goal of this subsection is to explain that Dish does a safe
%%   but effective analysis to identify which parts of the scripts to
%%   parallelize.}

@@

\noindent
Notice that the structure of a shell program can be used to identify
components that can be executed concurrently, and barriers that
enforce synchronization. More precisely, \ttt{cat f1 f2} and \ttt{grep
  ``foo''} would already execute concurrently in the standard shell,
while \ttt{sort f3} would wait for their completion before being
executed. Both \ttt{cat f1 f2 | grep ``foo'' > f3} and \ttt{sort f3}
are maximal distributable regions, in the sense that they cannot be
extended beyond \ttt{\&\&}.

Concretely, a distributable region cannot permeate through barrier
constructs (e.g. \ttt{;}, \ttt{\&\&}, and function definitions). On
the other hand, the pipe operator \ttt{|} and the parallel composition
operator \ttt{\&} enable parallelization, and extend distributable
regions. Intuitively, distributed regions correspond to subsections of
the program that would be allowed to execute concurrently by different
processes in the POSIX standard~\cite{posix}.

\tr{I don't know whether I should mention the following: While these
  control flow constructs can be parallelized and there has been
  research on it, it is orthogonal to our work, and can be
  incorporated as future work.}

@@

Dish performs a depth first search pass on the AST of the given shell
program, extending the distributable regions from the bottom up,
translating their independent components to dataflow graph
nodes, until a barrier construct is reached. Note that barrier
constructs can exist in a dataflow graph, as long as they don't
extend beyong a single node. An illustrative example follows:

@@

In the above example, while \ttt{sort f1 > g1}, \ttt{sort f2 > g2},
and \ttt{sort -m g1 g2} are all in different dataflow graphs, the
dataflow graph that corresponds to the whole program, has two nodes,
\ttt{sort f1 > g1 \&\& sort f2 > g2} and \ttt{sort -m g1 g2}.

In order to identify possible opportunities for parallelism, the
translation pass also identifies the category of each node, together
with its input and output, using records written by developers using
the language that was described in \Cref{ext}. Some interesting cases
of the analysis algorithm are shown in pseudocode below:

@@

Whenever the algorithm encounters a pipe node, it recursively
translates its subcomponents, and then merges them in a dataflow graph
by connecting the output of the first with the input of the second,
\etc. Before the different dataflow graphs are connected, the
algorithm checks that at most one node in the graph writes at every
file, and at most one node reads from a file that another node writes
to. This is important, in order to avoid inconsistent behaviour after
distributing nodes in the dataflow graph (i.e. by introducing to
concurrent reads and writes to the same file when
parallelizing). Whenever the algorithm encounter an \ttt{\&\&} node,
it recursively translates its subcomponents (similarly to the pipe
node), but then does not merge them in a large dataflow
graph. Finally, whenever the algorithm encounters a command, it first
checks whether it is safe to parallelize it. A command is considered
to be safe, if it does not make any modification to the shell state
(e.g. changing the current directory, modifying environment
variables), and if it does not read from any \todo{read-once} file,
like a fifo pipe. If the command is considered safe, its category is
identified and a node containing that command is created and added to
a singleton dataflow graph. If the command is considered unsafe, it is
still added to the dataflow graph, but cannot be further parallelized
by the optimizer.

An important consideration is that due to the highly dynamic nature of
the shell, the safety properties that are mentioned above cannot be
soundly checked statically. For example in the first case, two
unexpaned strings could refer to the same file, but the translation
pass cannot infer that.

@@

\heading{Limitations}
Our is implementation is limited in many ways, so as to succeed in proving the key hypothesis---that scaling out shell pipelines can be automated and correct\TODO{with respect to some assumptions}--- despite \sys being a research prototype.

@@

\sys is conservative in the shell subsets that it handles, in order to
not introduce unsafe behaviours. The point of this work was not to be
able to handle a complete distributable subset of the shell, but
rather use significant part of it to demonstrate performance benefits.

Shell is highly dynamic, thus making it impossible to design sound
meaningful analyses to determine distributable regions and maximal
parallelization. In order to address this particularity, \sys could be
tighly integrated with a posix compliant shell
interpreter~\cite{smoosh:20} so that dynamic information can be
acquired selectively to aid the analysis and optimization process.

\heading{Future Work}
% \label{limitation}
% \item No cycles (multiple commands writing and reading from the same file)
There are several worthwhile directions for future work.  One
direction would be an extension of the distributability analysis of
\sys, in order to guide it using dynamic information. This could
enable a formal study of the analysis~\sx{impl} to show that it
returns regions that are safe to distribute.  Recent work on
formalizing the semantics of the POSIX shell~\cite{smoosh:20} make
such an attempt possible.
